Mining of Keystroke and Mouse Dynamics to Enhance Student Engagement with Programming Assignments

### 2018-04-18 paper.tex
Other changes

In order to promote engagement, sometimes instruction is designed with the
objective of creating learning activities that are challenging to students, but
not much as to frustrate them. Engagement is not only achieved through the
design of single learning activities but also on the appropiate selection and
sequence in which they are presented.
For this, experienced instructors gauge a student\'s affective state and then
assign an activity with the appropriate level of difficulty. However,
recognition of emotions is a difficult task, even
for humans. Instructors often use their social intelligence to recognize
students\'' affective states. In class, an instructor habitually reads the
faces of students in the classroom to see if they are confused, bored or
engaged; then decides what to do next. Interactive Learning Environments (ILEs)
can be enhanced if they can offer an
automatic perception of emotions. The recognition and simulation of human
affects are becoming important fields of study, and many researchers have
demonstrated that affect-aware computers can provide a better performance in assisting
humans \cite{picard2001toward}. When ILEs embrace these methodologies and techniques a
problem arises. Sensors must be used to gather data, in order to perform the
recognition of users\' affective states.
Some sensors rely on physiological readings
and must be in physical contact with students.
These sensors can be considered
intrusive or invasive, and can disrupt the student\'s learning experience
\cite{zhai2008stress, sidney2005integrating, arroyo2009emotion}.
 Other sensors such as video cameras can also be considered invasive, since they always transmit the
 identity, appearance and behavior along with emotional data \cite{picard2001toward}.

In programming courses, students solve programming problems by implementing
their proposed solution in a particular language; in order to do this, they
must write the code,
some times compile it, to finally run it to see how it works. This cycle is
often repeated many times, and studens can change their affective state
accordingly.
Data can be collected, while students are working on their assignments.
All the dynamics of actually writing the code using the keyboard for instance
the syntax errors they received from the compiler or interpreter, the
corrections they tried, the elapsed time of each step, the number of attempts,
compiling errors, exceptions, among others. Data mining of all these data could
bring us a better understanding of the learning process, and also a better
model to both the selection of the appropriate content and to evaluate the
learner\'s competencies. But the problem that remains is to how to recognize
the student's affective state so that it can be used by an adaptive system.
% to (select appropriate content| give grades| whatever ) - JJ
% I highlited the Data Ming task for the track, and that there are other options
% for Data Analytics
In this work, a method for the recognition of affective states through the
mining of keystroke and mouse dynamics data is proposed.
The hypothesis is that students can vary the dynamic of keystrokes according
to their affective state when programming. A correlation of the
user\'s interactions (keyboard and mouse) with an emotional state, has been
established in the preliminary work by Zimmermann
et al. \cite{zimmermann2003affective}, but the work experiments where
conducted by asking users to write a predefined message, not programming
execercises.
% since Bosch or whatever has already proved that this is possible (or
% whatever) --> Establish the state of the art and how you plan to
% advance from it - JJ
In order to test the correlation of keyboard and mouse dynamics and an
affective state when programming, an interactive web-based
programming platform has been proposed \cite{ijcci17}.
The platform is designed to collect keyboard and mouse data from the students\' interaction for data analysis. The proposed method has three main components: a JavaScript library to capture keyboard and mouse dynamics from a browser-based editor, a pre-processing step whose output is a feature vector that is sent to a third component, the classification algorithm.

@@

The goal of this work is to propose an affective
recognition method based on the sensory data provided by the keyboard and mouse
dynamics generated by a learner as he/she types a programming assignment. A brief explanation of the
whole process is explained next, details come later. The process starts when the
learner begins to type a program in a browser-based editor. As she types or
moves the mouse the dynamics are recorded. When the learner submits the code to
evaluation, the request includes the sensory data along with the code and
information about the session. In the server, the code is evaluated in an
external virtual machine that provides a sandbox to prevent malicious or
erroneous code to halt the server. When the result is ready, it is recorded
along with the sensory data and sent to the preprocessing module. The output is
a feature vector, ready for classification. A previously trained classifier is
responsible for the classification, which outputs the predicted affective state.
The method does not consider other sensory data, but it could be integrated with
other sensory inputs in a multi-modal approach.  Each of the steps is explained
in detail next.

@@

\subsubsection{Capturing the Keystroke and Mouse Data}
While a student is trying
to solve an assignment, a script coded in JavaScript is running in the background,
which captures every keystroke, mouse movement and mouse button press. Each
capture of these events records a time-stamp in milliseconds (using the method
getTime() of JavaScript\'s built-in class Date) that describes when the event
occurred. If the event is a keystroke, the script captures what key was
specifically pressed, and what type of event occurred, it can be either a key-
down or a key-up event. If it is an event related to a mouse button press, the
key code of that button is recorded, as well as the type of event occurred again
key-down or key-up. Finally, if the event was a mouse movement, the mouse
coordinates inside of the web browser are recorded. The script monitors the
mouse position every 100 milliseconds, and only if the position has changed, it
records the new position. Each time a learner tries to evaluate a program, all
the data generated is sent to the server along with the code. When the result of
the execution is returned, all records are cleared and the process starts again.
There is no problem if the user leaves for a long period of time, because no
event will be triggered. If a user copies and then pastes the code from another
source, this will be recorded. There is a limitation, only the browser-editor
must be used; this could be a problem for more advanced programmers needing
specialized editors.  On the other hand a browser-based editor with the
corresponding remote execution, does not require the installation of
interpreters or compilers in learner’s computer. The code could even be written
in a mobile device or any web-enabled device. Programming assignments are evaluated using unit
tests; the results of the evaluation are shown to users. If all tests the
program is considered to be correct. The source code for the Protoboard web
based learning environment including the KD and MD functions are open source and
available as Github repositories at \url{https://github.com/amherag/keyboard-mouse-dynamics}; the code for the
sandbox is also available in \url{https://github.com/mariosky/sandbox}.

>>>>

In order to promote engagement, sometimes instruction is designed with the
objective of creating learning activities that are challenging to students, but
not much as to frustrate them. Engagement is not only achieved through the
design of single learning activities but also on the appropiate selection and
sequence in which they are presented.
For this, experienced instructors gauge a student\'s affective state and then
assign an activity with the appropriate level of difficulty. However,
recognition of emotions is a difficult task, even
for humans. Instructors often use their social intelligence to recognize
students\'' affective states. In class, an instructor habitually reads the
faces of students in the classroom to see if they are confused, bored or
engaged; then decides what to do next. Interactive Learning Environments (ILEs)
can be enhanced if they can offer an
automatic perception of emotions. The recognition and simulation of human
affects are becoming important fields of study, and many researchers have
demonstrated that affect-aware computers can provide a better performance in
assisting humans \cite{picard2001toward}. When ILEs embrace these methodologies
and techniques a
problem arises. Sensors must be used to gather data, in order to perform the
recognition of users\' affective states.
Some sensors rely on physiological readings
and must be in physical contact with students.
These sensors can be considered
intrusive or invasive, and can disrupt the student\'s learning experience
\cite{zhai2008stress, sidney2005integrating, arroyo2009emotion}.
Other sensors such as video cameras can also be considered invasive, since
they always transmit the user's identity, appearance and behavior along with
the emotional data \cite{picard2001toward}.

In programming courses, students solve programming problems by implementing
their solution in a particular language; in order to do this, they
must type the code, some times compile it, and finally run it to see how or if
it works. This cycle can be repeated many times, as they are in this cycle,
studens often change their affective state according to their performence.
Data can be collected while students are working on their assignments.
All the dynamics of actually typing the code and for instance
the syntax errors they received from the compiler or interpreter, the
corrections they tried, the elapsed time of each step, the number of attempts,
compiling errors, exceptions, among others. Data mining of all this data could
bring us a better understanding of the learning process, and also a better
model to both the selection of the appropriate content and to evaluate the
learner\'s competencies. But the problem that remains is to how to recognize
the student's affective state so that it can be used by an adaptive system.
% to (select appropriate content| give grades| whatever ) - JJ
% I highlited the Data Ming task for the track, and that there are other options
% for Data Analytics
In this work, a method for the recognition of affective states through the
mining of keystroke and mouse dynamics data is proposed.
The hypothesis is that students can vary the dynamic of keystrokes according
to their affective state when programming. A correlation of the
user\'s interactions (keyboard and mouse) with an emotional state, has been
established in the preliminary work by Zimmermann
et al. \cite{zimmermann2003affective}, but the work experiments where
conducted by asking users to write a predefined message, not programming
execercises.
% since Bosch or whatever has already proved that this is possible (or
% whatever) --> Establish the state of the art and how you plan to
% advance from it - JJ
In order to test the correlation of keyboard and mouse dynamics and an
affective state when programming, an interactive web-based
programming platform has been proposed \cite{ijcci17}.
The platform is designed to collect keyboard and mouse data from the students\' interaction for data analysis. The proposed method has three main components: a JavaScript library to capture keyboard and mouse dynamics from a browser-based editor, a pre-processing step whose output is a feature vector that is sent to a third component, the classification algorithm.

@@

The goal of this work is to propose an affective
recognition method based on the sensory data provided by the keyboard and mouse
dynamics generated by a learner as he or she types a programming assignment.
A detailed explanation of the process outlined in a previous work \cite{ijcci17} is explained next.
The process starts when the
learner begins to type a program in a browser-based editor. As she types or
moves the mouse all the data of the dynamics is recorded on the client.
When the learner submits the code to evaluation, the HTTP request includes,
in a JSON string, all the sensory data along with the code and
information about the current session.
In the web server, the code is evaluated by an
external container that provides a virtual sandbox in which to execute the
program. The sandbox is used to prevent malicious or
erroneous code to halt the server. When a container is halted, it is simply
removed and a new container is created.
When the result is ready, it is recorded
along with the sensory data and sent to the preprocessing module that it will
be explained next. The output is a feature vector, ready for classification. A
previously trained classifier is responsible for the classification, which
outputs the predicted affective state.

Currently, the method does not consider other sensory data,
but it could be integrated with other sensory inputs in a multi-modal
approach.  Each of the steps is explained in detail next.

@@

\subsubsection{Capturing the Keystroke and Mouse Data}
As it was explained before, as a student is trying to solve an assignment,
a script coded in JavaScript runs in the background,
capturing every keystroke, mouse movement and mouse button pressed. Each record
of these events includes a time-stamp in milliseconds (using the method
getTime() of JavaScript\'s built-in class Date) that describes when the event
occurred. If the event is a keystroke, the script captures what key was
specifically pressed, and what type of event occurred, it can be either a key-
down or a key-up event. If it is an event related to a mouse button press, the
key code of that button is recorded, as well as the type of event occurred again
key-down or key-up. Finally, if the event was a mouse movement, the mouse
coordinates inside of the web browser are recorded. The script monitors the
mouse position every 100 milliseconds, and only if the position has changed, it
records the new position. Each time a learner tries to evaluate a program, all
the data generated is sent to the server along with the code. When the result of
the execution is returned, all records are cleared and the process starts again.
There is no problem if the user leaves for a long period of time, because no
event will be triggered. If a user copies and then pastes the code from another
source, this will be recorded. There is a limitation, only the browser-editor
must be used; this could be a problem for more advanced programmers needing
specialized editors.  On the other hand a browser-based editor with the
corresponding remote execution, does not require the installation of
interpreters or compilers in learner’s computer. The code could even be written
in a mobile device or any web-enabled device. Programming assignments are
evaluated using unit tests; the results of the evaluation are shown to users.
If all tests the program is considered to be correct. The source code for the
Protoboard web based learning environment including the KD and MD functions are
open source and available as Github repositories
at\url{https://github.com/amherag/keyboard-mouse-dynamics}; the code for the
sandbox is also available in \url{https://github.com/mariosky/sandbox}.
